---
title: "Satellites and Bedrock - Part II"
description: |
  Looking at the ways in which satellite images from Sentinel could be used to examine underlying bedrock types. In this second of a three part series, we look at ways of accessing satellite data from the Sentinel satellites.
author:
  - name: Eugene
    url: https://www.fizzics.ie
date: 2023-01-19
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Geological Survey Ireland is producing wonderful [data resources](https://www.gsi.ie/en-ie/data-and-maps/Pages/default.aspx), impossible to resist when it comes to doing some analysis. In this post, welook at [bedrock data](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), what you would see after sweeping away all the soil and loose covering from the ground. Bedrock is stuff like sandstone, limestone, igneous rock.... 

I was interested to see how overlying vegetation depends on the bedrock; specifically if the colours from satellite images taken at different times of year could be used to take a guess at what lies beneath. 

This requires two suites of information, the bedrock data from GSI and the satellite data from [Copernicus-Sentinel](https://scihub.copernicus.eu/). The first we'll download directly and store locally from [this resource](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), the second we'll access using the R package, [sen2r](https://github.com/ranghetti/sen2r) (but we'll also end up downloading and storing image files locally, but `sen2R` will help scope them out).

This was originally written as one post, but then it got so long I decided to split it into three parts:

- [Part I: getting the bedrock data from GSI](https://www.fizzics.ie/posts/2023-01-17-satellites-and-bedrock/)  
- **Part II: getting the satellite data from Sentinel  **
- Part III: putting the first two together  


```{r libraries-theme}
library(tidyverse)
library(sf)
library(sen2r)
library(raster)
library(RStoolbox)
library(exactextractr)
# library(terra)
library(showtext)
library(gt)

font_add_google(name = "Gloria Hallelujah", family = "my_font")
showtext_auto()
theme_clean <- function() {
  theme_minimal(base_family = "my_font") +
    theme(panel.grid.minor = element_blank(),
          text = element_text(size = 16, family = "my_font"),
          plot.background = element_rect(fill = "white", color = NA),
          axis.text = element_text(size = 16),
          axis.title = element_text(face = "bold", size = 14),
          strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0),
          strip.background = element_rect(fill = "grey80", color = NA),
          legend.text = element_text(size = 16, family = "my_font"))
}
```

This post is largely inspired by the work of [Ewa Grabska](https://github.com/WhyR2020/workshops/blob/master/satellite/whyr_satellite.R) and a [WhyR presentation she made in 2020](https://www.youtube.com/watch?v=k1K6nqgtRL8). She has a much clearer and in-depth view of the topic than I do.

Getting access to [Sentinel satellite](https://scihub.copernicus.eu/) data requires making an account at [scihub](https://scihub.copernicus.eu/dhus/#/self-registration). It's free and open access. And once you have your login details you can do most of the leg work without leaving rstudio thanks to the `sen2r` package. As we'll see, the only time we'll leave rstudio is to unzip a downloaded file.