---
title: "Satellites and Bedrock - Part III"
description: |
    Looking at the ways in which satellite images from Sentinel could be used to examine underlying bedrock types. In this third of a three part series, we look at ways of stacking the satellite data and then squaring that off with what we've seen from bedrock.
author:
  - name: Eugene
    url: https://www.fizzics.ie
date: 2023-01-21
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Geological Survey Ireland is producing wonderful [data resources](https://www.gsi.ie/en-ie/data-and-maps/Pages/default.aspx), impossible to resist when it comes to doing some analysis. In this post, welook at [bedrock data](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), what you would see after sweeping away all the soil and loose covering from the ground. Bedrock is stuff like sandstone, limestone, igneous rock.... 

I was interested to see how overlying vegetation depends on the bedrock; specifically if the colours from satellite images taken at different times of year could be used to take a guess at what lies beneath. 

This requires two suites of information, the bedrock data from GSI and the satellite data from [Copernicus-Sentinel](https://scihub.copernicus.eu/). The first we'll download directly and store locally from [this resource](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), the second we'll access using the R package, [sen2r](https://github.com/ranghetti/sen2r) (but we'll also end up downloading and storing image files locally, but `sen2R` will help scope them out).

This was originally written as one post, but then it got so long I decided to split it into three parts:

- [Part I: getting the bedrock data from GSI](https://www.fizzics.ie/posts/2023-01-17-satellites-and-bedrock/)  
- [Part II: getting the satellite data from Sentinel](https://www.fizzics.ie/posts/2023-01-19-satellites-and-bedrock-part-ii/)
- **Part III: putting the first two together  **


```{r libraries-theme}
library(tidyverse)
library(sf)
library(sen2r)
library(getSpatialData)
library(raster)
library(RStoolbox)
library(exactextractr)
# library(terra)
library(showtext)
library(gt)
library(ggokabeito)

font_add_google(name = "Gloria Hallelujah", family = "my_font")
showtext_auto()
theme_clean <- function() {
  theme_minimal(base_family = "my_font") +
    theme(panel.grid.minor = element_blank(),
          text = element_text(size = 16, family = "my_font"),
          plot.background = element_rect(fill = "white", color = NA),
          axis.text = element_text(size = 16),
          axis.title = element_text(face = "bold", size = 14),
          strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0),
          strip.background = element_rect(fill = "grey80", color = NA),
          legend.text = element_text(size = 16, family = "my_font"))
}
```

```{r bedrock}
ref <- st_read("../2023-01-17-satellites-and-bedrock/data/Shapefiles/Bedrock_Polygons_ITM_2018.shp",
               quiet = TRUE)
my_proj <- "+proj=utm +zone=29 +datum=WGS84 +units=m +no_defs"
ref <- st_transform(ref, crs = my_proj)
```


```{r}
prefix <- "../../../"

my_file_red_june <- glue::glue("{prefix}data/spring-2022/T29UNA_20220604T114401_B02.jp2")
my_file_green_june <- glue::glue("{prefix}data/spring-2022/T29UNA_20220604T114401_B03.jp2")
my_file_blue_june <- glue::glue("{prefix}data/spring-2022/T29UNA_20220604T114401_B04.jp2")
my_file_nir_june <- glue::glue("{prefix}data/spring-2022/T29UNA_20220604T114401_B08.jp2")

my_file_red_aug <- glue::glue("{prefix}data/summer-2022/T29UNA_20220811T115409_B02.jp2")
my_file_green_aug <- glue::glue("{prefix}data/summer-2022/T29UNA_20220811T115409_B03.jp2")
my_file_blue_aug <- glue::glue("{prefix}data/summer-2022/T29UNA_20220811T115409_B04.jp2")
my_file_nir_aug <- glue::glue("{prefix}data/summer-2022/T29UNA_20220811T115409_B08.jp2")

my_file_red_dec <- glue::glue("{prefix}data/december-2022/T29UNA_20221204T115451_B02.jp2")
my_file_green_dec <- glue::glue("{prefix}data/december-2022/T29UNA_20221204T115451_B03.jp2")
my_file_blue_dec <- glue::glue("{prefix}data/december-2022/T29UNA_20221204T115451_B04.jp2")
my_file_nir_dec <- glue::glue("{prefix}data/december-2022/T29UNA_20221204T115451_B08.jp2")
```

