---
title: "Satellites and Bedrock"
description: |
  Looking at the ways in which satellite images from Sentinel could be used to examine underlying bedrock types.
author:
  - name: Eugene
    url: https://www.fizzics.ie
date: 2023-01-17
categories: [R, Donegal, Maps]
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
```


Geological Survey Ireland is producing wonderful [data resources](https://www.gsi.ie/en-ie/data-and-maps/Pages/default.aspx), impossible to resist when it comes to doing some analysis. In this post, welook at [bedrock data](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), what you would see after sweeping away all the soil and loose covering from the ground. Bedrock is stuff like sandstone, limestone, igneous rock.... 

I was interested to see how overlying vegetation depends on the bedrock; specifically if the colours from satellite images taken at different times of year could be used to take a guess at what lies beneath. 

This requires two suites of information, the bedrock data from GSI and the satellite data from [Copernicus-Sentinel](https://scihub.copernicus.eu/). The first we'll download directly and store locally from [this resource](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), the second we'll access using the R package, [sen2r](https://github.com/ranghetti/sen2r) (but we'll also end up downloading and storing image files locally, but `sen2R` will help scope them out).

The bedrock file arrives in the form of a shapefile. It contains bunch of attributes: location, decription, area, as well as the polygon shapes themselves. And some fields I'm at a loss to understand (M, Y, K, C anyone?).

```{r libraries-theme}
library(tidyverse)
library(sf)
library(sen2r)
library(raster)
library(RStoolbox)
library(exactextractr)
# library(terra)
library(showtext)

font_add_google(name = "Gloria Hallelujah", family = "my_font")
showtext_auto()
theme_clean <- function() {
  theme_minimal(base_family = "my_font") +
    theme(panel.grid.minor = element_blank(),
          text = element_text(size = 16, family = "my_font"),
          plot.background = element_rect(fill = "white", color = NA),
          axis.text = element_text(size = 24),
          axis.title = element_text(face = "bold", size = 28),
          strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0),
          strip.background = element_rect(fill = "grey80", color = NA),
          legend.text = element_text(size = 16))
}
```



```{r bedrock, echo=TRUE}
ref <- suppressMessages(st_read("data/Shapefiles/Bedrock_Polygons_ITM_2018.shp"))
my_proj <- "+proj=utm +zone=29 +datum=WGS84 +units=m +no_defs"
ref <- st_transform(ref, crs = my_proj)
```

```{r}
ref$geometry
```


```{r}
ref |> st_drop_geometry() |> head() |> gt::gt()
```



