---
title: "Satellites and Bedrock - Part I"
description: |
  Looking at the ways in which satellite images from Sentinel could be used to examine underlying bedrock types. In this first of a three part series, we look at ways of accessing bedrock data from GSI.
author:
  - name: Eugene
    url: https://www.fizzics.ie
date: 2023-01-18
categories: [R, Donegal, Maps]
output:
  distill::distill_article:
    self_contained: false
---

```{css, echo=FALSE}
img {
  max-width:150%;
  height: auto;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
```


Geological Survey Ireland is producing wonderful [data resources](https://www.gsi.ie/en-ie/data-and-maps/Pages/default.aspx), impossible to resist when it comes to doing some analysis. In this post, welook at [bedrock data](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), what you would see after sweeping away all the soil and loose covering from the ground. Bedrock is stuff like sandstone, limestone, igneous rock.... 

I was interested to see how overlying vegetation depends on the bedrock; specifically if the colours from satellite images taken at different times of year could be used to take a guess at what lies beneath. 

This requires two suites of information, the bedrock data from GSI and the satellite data from [Copernicus-Sentinel](https://scihub.copernicus.eu/). The first we'll download directly and store locally from [this resource](https://www.gsi.ie/en-ie/data-and-maps/Pages/Bedrock.aspx#100k), the second we'll access using the R package, [sen2r](https://github.com/ranghetti/sen2r) (but we'll also end up downloading and storing image files locally, but `sen2R` will help scope them out).

This was originally written as one post, but then it got so long I decided to split it into three parts:

- **Part I: getting the bedrock data from GSI  **
- Part II: getting the satellite data from Sentinel  
- Part III: putting the first two together  


```{r libraries-theme}
library(tidyverse)
library(sf)
library(sen2r)
library(raster)
library(RStoolbox)
library(exactextractr)
# library(terra)
library(showtext)
library(gt)
library(ggokabeito)

font_add_google(name = "Gloria Hallelujah", family = "my_font")
showtext_auto()
theme_clean <- function() {
  theme_minimal(base_family = "my_font") +
    theme(panel.grid.minor = element_blank(),
          text = element_text(size = 16, family = "my_font"),
          plot.background = element_rect(fill = "white", color = NA),
          axis.text = element_text(size = 16),
          axis.title = element_text(face = "bold", size = 14),
          strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0),
          strip.background = element_rect(fill = "grey80", color = NA),
          legend.text = element_text(size = 16, family = "my_font"))
}
```


```{r bedrock, echo=TRUE}
ref <- st_read("data/Shapefiles/Bedrock_Polygons_ITM_2018.shp", quiet = TRUE)
my_proj <- "+proj=utm +zone=29 +datum=WGS84 +units=m +no_defs"
ref <- st_transform(ref, crs = my_proj)
```

The bedrock file arrives in the form of a shapefile. It contains bunch of attributes: location, decription, area, as well as the polygon shapes themselves. And some attributes I'm at a loss to understand (M, Y, K, C anyone?). We end up with a total of `r dim(ref)[1] |> format(big.mark=",")` fields covering the Republic of Ireland. 

Below we show the geometry for this shapefile:


```{r}
ref$geometry
```

And here is a table showing the attributes for a sample of a few fields. Note, we've done a wee text-trawl through the description attribute to try and categorise the rocks in to a manageable number of types (quartz, sandstone, limestone, schist, shale, granite, or other).

```{r table, layout="l-screen-inset shaded"}
ref |>
  mutate(
    class = case_when(
      str_detect(DESCRIPT, "[Gg]ranite") ~ "granite",
      str_detect(DESCRIPT, "[Qq]uar[tz][tz]") ~ "quartz",
      str_detect(DESCRIPT, "[Ss]andstone") ~ "sandstone",
      str_detect(DESCRIPT, "[Ss]chist") ~ "schist",
      str_detect(DESCRIPT, "[Ll]imestone") ~ "limestone",
      str_detect(DESCRIPT, "[Ss]hale") ~ "shale",
      TRUE ~ "other"
    ),
  .before = NEWCODE
  ) |> 
  st_drop_geometry() |> 
  dplyr::select(-c(OBJECTID_2, OBJECTID, SHEETNO, STRATCODE, LITHCODE, SHAPE_LEN)) |> 
  slice_sample(n=10) |> 
  gt::gt() |> 
  tab_options(table.font.size = px(16)) |> 
  fmt_number(columns = SHAPE_AREA,
             rows = SHAPE_AREA > 1000,
             scale_by = 1/1000,
             decimals = 0,
             pattern = "{x}K") |> 
  fmt_number(columns = SHAPE_AREA,
             rows = SHAPE_AREA < 1000,
             scale_by = 1,
             decimals = 2,
             pattern = "{x}")
```

```{r bedock-map}
my_extent <- c(xmin=540000, xmax=585000, ymin=6050000, ymax=6090000)
my_extent <- c(xmin=510000, xmax=635000, ymin=6030000, ymax=6150000)


# ref <- st_transform(ref, crs = st_crs(s1))
ref1 <- ref |> 
  st_crop(extent(my_extent)) |> 
  mutate(
    class = case_when(
      str_detect(DESCRIPT, "[Gg]ranite") ~ "granite",
      str_detect(DESCRIPT, "[Qq]uar[tz][tz]") ~ "quartz",
      str_detect(DESCRIPT, "[Ss]andstone") ~ "sandstone",
      str_detect(DESCRIPT, "[Ss]chist") ~ "schist",
      str_detect(DESCRIPT, "[Ll]imestone") ~ "limestone",
      str_detect(DESCRIPT, "[Ss]hale") ~ "shale",
      TRUE ~ "other"
    ),
  .before = NEWCODE)
#  |> filter(class != "other")
```

Plotting this, and just looking at the NorthWest, we get:

```{r map, layout="l-screen-inset shaded"}
ref1 |> 
  ggplot(aes(fill = class)) +
  geom_sf() + 
  guides(fill=guide_legend(nrow=1,byrow=TRUE)) +
  scale_fill_okabe_ito() +
  theme_clean() + 
  theme(legend.position = "bottom",
        legend.title = element_blank())

```



