---
title: "University Points"
description: |
  How do entry level points vary by university. And a chance to play with geomtextpath.
author:
  - name: Eugene
    url: https://fizzics.netliffy.app
date: 2021-12-15
categories: [Academia, R]
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We looked at [university choices](https://fizzics.netlify.app/posts/2021-12-06-university-choices/) last week, and the way in which different third level courses have different entry requirements. Using the same data, I thought it might be fun to see which were the most sought after universities. It's also an excuse to play with [geomtextpath](https://github.com/AllanCameron/geomtextpath), a package I haven't used before but that looks like it has some nifty features for `ggplot2` work.

We get the data as in the [university choices](https://fizzics.netlify.app/posts/2021-12-06-university-choices/) post. 



```{r libraries}
# http://www2.cao.ie/points/deg03.htm#trd

library(tidyverse)
library(pdftools)
library(ggokabeito)   # Colorblind-friendly color palette
library(showtext)
library(geomtextpath)
```


```{r graphics}
font_add("Fuzzy Bubbles", regular = "fonts/FuzzyBubbles-Regular.ttf")
showtext_auto()

theme_clean <- function() {
  theme_minimal(base_family = "Fuzzy Bubbles") +
    theme(panel.grid.minor = element_blank(),
          text = element_text(size = 28, family = "Fuzzy Bubbles"),
          plot.background = element_rect(fill = "white", color = NA),
          axis.text = element_text(size = 32),
          axis.title = element_text(face = "bold", size = 36),
          strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0),
          strip.background = element_rect(fill = "grey80", color = NA),
          legend.text = element_text(size = 24))
}
```


```{r data_access, cache = T, echo = F}
years <- c("07", "08", "09", "10", "13", 14:19 )
colleges <- tribble(~college, ~name,
                    "BN", "Blanchestown",
                    "CK", "UCC",
                    "CW", "Carlow IT",
                    "DC", "DCU",
                    "DK", "Dundalk IT",
                    "DN", "UCD",
                    "GY", "Univ Galway",
                    "LC", "Limerick IT",
                    "LM", "Univ Limerick",
                    "MH", "Maynooth",
                    "SG", "Sligo IT",
                    "TL", "Tralee IT",
                    "TR", "Trinity",
                    "TU", "TUD",
                    "WD", "W'ford IT")
cao_points_year <- function(year) {
  cao_pdf <- glue::glue("http://www2.cao.ie/points/lvl8_{year}.pdf")
  z <- pdf_text(cao_pdf) %>% 
    str_split("\n") %>% 
    unlist()
  z <- z[!str_detect(z, "^ ") & z != "" & !str_detect(z, "Course Code") & str_count(z, "  +") == 3] %>% # gets rid of non-data rows
    str_split("  +") %>% # splits rows based on runs of several spaces
    unlist() %>% 
    str_remove("#") %>% 
    str_remove("\\*") # deletes some annoying characters
  z <- tibble(year = glue::glue("20{year}"), 
              code = z[c(T, F, F, F)], 
              course = z[c(F, T, F, F)], 
              final = z[c(F, F, T, F)], 
              medium = z[c(F, F, F, T)]) 
  # original list made bunches of four elements that together described a course. This tibble winds them up to one row
  z
}

z <- map_df(years, cao_points_year) %>% 
  mutate(year = as.numeric(year))
```


```{r data_access_xl, cache = T, echo = F}
z21 <- readxl::read_excel("data/CAOPointsCharts2021.xlsx", 
                        sheet = "EOS_2021", skip = 10) %>% 
  janitor::clean_names() %>% 
  filter(course_level == 8) %>% 
  mutate(year = 2021) %>% 
  select(year,
         code = course_code, 
         course = course_title,
         final = eos_points,
         medium = eos_midpoints) %>% 
  mutate(year = 2021,
         medium = as.character(medium))
z20 <- readxl::read_excel("data/CAOPointsCharts2020.xlsx", 
                        sheet = "PointsCharts2020_V2", skip = 9) %>% 
  janitor::clean_names() %>% 
  filter(level == 8) %>% 
  mutate(year = 2020) %>%  
  select(year,
         code = course_code2, 
         course = course_title,
         final = eos,
         medium = eos_mid_point) %>% 
  mutate(year = 2020)
```


```{r data_access_combo, cache = T, echo = F}
z <- bind_rows(z, z20, z21) %>% 
  mutate(final = as.numeric(final), medium = as.numeric(medium)) %>%
  filter(!is.na(final), !is.na(medium)) %>% 
  mutate(college = str_sub(code, 1, 2)) %>% 
  left_join(colleges)
```

This gives a data frame something like this:

```{r table, echo=FALSE}
DT::datatable(z %>% select(-name))
```


```{r figure_out, fig.height=6, echo = T}
z %>% filter(college %in% c("TR", "TU", "MH")) %>% 
  ggplot(aes(x = final, colour = name)) +
  geom_textpath(aes(label = name), stat = "density",
                size = 12, fontface = 2, hjust = 0.2, vjust = 0.3,
                show.legend = F, linewidth = 2, family = "Fuzzy Bubbles") +
  scale_color_okabe_ito() +
  theme_clean() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank()) +
  coord_cartesian(xlim = c(200, 700))
```

Looks like a clear pecking order, with Trinity College Dublin top of the pile, colleges like Maynooth University in the middle, and the newer colleges like Technological University Dublin catching up.
