---
title: "Schizophrenia Genes"
description: |
  A karyogram of the genes featured in GWAS of schizophrenia.
author:
  - name: Eugene
    url: https://fizzics.netlify.app
date: 04-26-2021
categories: [Bioinformatics]
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


We've been working with the genetics of schizophrenia, looking at how the expression of genes that play a role in the condition vary across brain regions and neurodevelopment stages. Now, call me old-fashioned, but I like to know how things are laid out, so in this case where these genes are situated. Time for a plot.

The genes in question come from a [GWAS study of the PGC](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5918692/). The images are build based on the [ggbio](https://www.bioconductor.org/packages/release/bioc/html/ggbio.html) and [karyoploteR](http://bioconductor.org/packages/release/bioc/html/karyoploteR.html) packages from bioconductor.

First, let's load up the required packages:

```{r libraries}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

if (!require("pacman")) install.packages("pacman")

pacman::p_load(annotables, tidyverse, ggbio,
               httr, readxl, ABAData, 
               GenomicRanges, karyoploteR, genes)

```

Next, we get a list of the genes from the PGC. These are taken from one of the supplementary tables from the [2018 paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5918692/) mentioned above. We only care about the ones that also feature in the [Allen Brain Atlas](https://portal.brain-map.org/), so we filter for those.

```{r gene_list, cache = T}
genes <- genes::pardinas() %>% 
  filter(genes %in% abagenes$hgnc_symbol) %>% 
  pull(genes)
```

From our gene list we build a data frame of gene information using Human build 38 (grch38 from the [annotables package](https://www.rdocumentation.org/packages/annotables/versions/0.1.1)). And then turn his into an appropriate GRanges object (thank you [GenomicRanges](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html)).

```{r GRanges, cache  = T}
gene_table <- grch38 %>% 
  dplyr::filter(symbol %in% genes) %>% 
  dplyr::filter(chr %in% c(1:22, "X", "Y")) %>% 
  dplyr::mutate(strand = ifelse(strand == 1, "+", "-"))

gene_ranges <- makeGRangesFromDataFrame(gene_table, keep.extra.columns = T)
seqlevelsStyle(gene_ranges) <- "UCSC"
```

Now for the plot. The `plotKaryotype()` function does a lqyout of 24 chromosomes with their cytobands, the `kpPlotMarkers()` adds in the gene labels. The label colours depend on strand; red for *+* and green for *-*. I wasn't too successful avoiding label overlaps, sorry about that. 

```{r karyogram}
kp <- plotKaryotype(genome="hg38")
kpPlotMarkers(kp, 
              data=gene_ranges, 
              labels=gene_ranges$symbol, 
              label.color = ifelse(gene_table$strand == "+", "darkred", "olivedrab4"),
              text.orientation = "horizontal",
              label.dist = 0.0001,
              r1=0.5, cex=0.6, adjust.label.position = T)
```

Our genes are pretty spread out, with no particular pattern of course. But I just feel I know them a little better now.