---
title: "Finding Postcodes using `arcgisgeocode`"
description: |
  Using the arcgiscoder package to find postal codes, making a shiny app using leaflet.
author:
  - name: Eugene
    url: https://www.fizzics.ie
date: 2024-06-25
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm old fashioned in a lot of ways, one of which is that I like to send postcards when on holidays. This means making sure they make their way to the intended recipients. And while I can usually remember an address, coming up with the corresponding postcode is at best hit-and-miss. There's no telling how many postcards down the years are still out there drifting through mail-sorting limbo.

Then I listened to a presentation by  [Josiah Parry]() and felt I had a solution to my problem. He introduced a package called `arcgisgeocoder` he's been working on, part of the `arcgis` system. One of its capabilities is to take a geographic location and feed back details about the address, including the postal code. Kind of what I need. Josiah went on to incorporate this into a `shiny` app using `leaflet`, something I'm going to copy here. Then, when you scroll around the map and click on the location, it gives you the address and postal code.

Let's see how this works. First, we'll need some packages:

```{r eval=FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(shiny)
library(bslib)
library(sf)
library(arcgisgeocode)
```

Then we create our `shiny` app itself. Note the function `reverse_geocode` from `arcgisgeocode`. This does all the heavy lifting. The rest is fairly standard map interaction; displaying a map and capturing the latitude and longitude of a point clicked.

```{r eval=FALSE}
ui <- page_fillable(
  card(
    card_title(textOutput("rev_result")),
    leafletOutput("map", width = "100%", height = "100%"),
  )
)

server <- function(input, output, session){
  observeEvent(input$map_click, {
    click <- input$map_click
    x <- click$lng
    y <- click$lat
    loc <- c(x, y)
    dput(loc)
    
    geocoded <- reverse_geocode(loc)
    
    output$rev_result <- renderText(
      glue::glue("{geocoded$long_label}\n{geocoded$postal}")
      )
    
    leafletProxy("map", data = st_geometry(geocoded)) |> 
      clearMarkers() |> 
      addMarkers()
  })
  
  output$map <- renderLeaflet({
    leaflet() |> 
      addProviderTiles(providers$Esri.WorldGrayCanvas) |> 
      setView(lat = 53.32041, lng = -6.23956, zoom = 14)
  })
}


shinyApp(ui, server)
```

Unfortunately, shinyapps.io can't handle the `arcgiscoder` package just yet, so I can't deploy this app. Yet. But I'll keep trying to find a way around this.


