---
title: "patchwork & gt tables"
description: |
  gt makes beautiful tables, and patchwork is a wonderful package for combining plots. This post shows a way to make them work together.
author:
  - name: Eugene
    url: https://www.fizzics.ie
date: 2022-02-22
categories: [Academia, R]
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r libraries, echo = F}
library(tidyverse)
library(gt)
library(patchwork)
library(palmerpenguins)
```


There are many options to make tables in R (`knitr::kable`, `formattable`, `flextable`....),  but `gt` is pretty much my favourite. It makes magnificent tables, and closely follows the philosophy, if not the syntax, of `ggplot`.   

One problem with `gt` tables, is that they don't sync well with `patchwork`. This is because `patchwork` needs grobs, and `gt` doesn't do that. For most types of tables, the `gridExtra::tableGrob` function would do the job, but not so for `gt` tables. This post presents a workaround (well, really a couple of fudges), so that we can get around this. We can use the beauty of `gt` tables, with all the flexibility of `patchwork` layouts.   

First off, I'd like all our tables to have the same look. So I'm going to create a common theme to apply to each of them. This is our first little hack here, as I haven't seen this done elsewhere.

```{r gt-theme}
my_theme <- function(data){
  data %>% 
    opt_table_font(font = google_font("Annie Use Your Telescope")) %>% 
    tab_options(column_labels.background.color = "#176940",
                table.font.size = px(16)
    )
}
```

Next, let's generate some tables. Let's use the familiar `penguins` dataset from the `palmerpengiuns` package, and make the following tables:

```{r three-tables}
table1 <- penguins %>% 
  group_by(species) %>% 
  summarise(weight = mean(body_mass_g, na.rm = T) %>% signif(3)/1000) %>% 
  ungroup() %>% 
  gt() %>% 
  my_theme()
table1

table2 <- penguins %>% 
  group_by(species) %>% 
  summarise(bill_length = mean(bill_length_mm, na.rm = T) %>% signif(3)) %>% 
  ungroup() %>%
  gt() %>% 
  my_theme()
table2

table3 <- penguins %>% 
  group_by(year, island) %>% 
  summarise(weight = mean(body_mass_g, na.rm = T) %>% signif(3)/1000) %>% 
  ungroup() %>%
  gt() %>% 
  my_theme()
table3  
```

Now for our second hack. We save each table in turn as a tmp file and then read it back in as a .png. This idea is from Johannes Enevoldsen given at https://github.com/rstudio/gt/issues/420. 

```{r grobs, include=FALSE}
# https://github.com/rstudio/gt/issues/420

tmp <- tempfile(fileext = '.png') #generate path to temp .png file
gtsave(table1, tmp) #save gt table as png
table_png1 <- png::readPNG(tmp, native = TRUE) # read tmp png file
gtsave(table2, tmp) #save gt table as png
table_png2 <- png::readPNG(tmp, native = TRUE) # read tmp png file
gtsave(table3, tmp) #save gt table as png
table_png3 <- png::readPNG(tmp, native = TRUE) # read tmp png file
```

And now the final hack. It seems like patchwork needs to kick off with an actual ggplot object, we can't just combine tables. So we create a blank plot and then make sure it's underneath everything when we apply the `patchwork::plot_layout`. 

```{r patches}
layout <- c(
  area(t=1, l=2, b=2, r=2),
  area(t=0, l=0, b=3, r=2),
  area(t=4, l=0, b=6, r=2),
  area(t=0, l=3, b=6, r=6)
)

(ggplot() + theme_void()) + table_png1 + table_png2 + table_png3 + 
  plot_layout(design = layout)
```



And there you have it. Of course, this is just a rudimentary layout, we haven't even begun to harness the full repetoire of either `gt` or `patchwork`. But it's a start.

